### 2015-06-23 stochastic/stochastic-recovery.Rmd
update stochastic approximation code/text

```{r}
set.seed(1)
n_spp = 100
n_loc = 1000
n_env = 3
n_gibbs = 1000

true_beta_vec = rnorm(
  choose(n_spp, 2), 
  sd = abs(rnorm(choose(n_spp, 2))) / 5
)

@@

true_env = matrix(rnorm(n_loc * n_env, 0), nrow = n_loc)

true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 0.5), nrow = n_env)

true_alpha = true_env %*% true_alpha_env
```

@@

```{r}
initial_learning_rate = 0.1
momentum = 0.9

pb <- progress_bar$new(
  format = "  Fitting [:bar] :percent eta: :eta",
  total = n_gibbs,
  clear = FALSE
)

for(i in 1:n_gibbs){
  pb$tick()

@@

  # Update learning rate
  learning_rate = initial_learning_rate * 1000 / (999 + i)

@@

  # Calculate the gradient with respect to alpha and beta
  delta_stats = y_stats - y_sim_stats
  beta_grad = (delta_stats - beta) / 10 / n_loc # Subtract beta as weight decay

  alpha_species_grad = diag(beta_grad) / n_loc
  diag(beta_grad) = 0
  alpha_env_grad = (y_env_stats - y_sim_env_stats - alpha_env)  / n_loc

  # Calculate parameter updates
  delta_beta = beta_grad * learning_rate + momentum * delta_beta
  delta_alpha_species = alpha_species_grad * learning_rate # no momentum
  delta_alpha_env = alpha_env_grad * learning_rate + momentum  * delta_alpha_env

@@

  if(i %% 10 == 0){
    plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
         pch = 16, cex = .5, col = "#00000050",
         xlab = "Estimated interaction strength",
         ylab = "\"True\" interaction strength",
         main = paste(
           "R^2 =", 
           signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
          )
    )
    abline(v = 0, h = 0)
  }
}
```

```{r}
pdf("poster-beta.pdf", width = 4, height = 5)
plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
     pch = 16, cex = .4, col = "#00000020",
     xlab = "Estimated interaction strength",
     ylab = "\"True\" interaction strength",
     main = paste0("Recoverd pairwise interactions", "\nN = ", choose(n_spp, 2)),
     las = 1,
     cex.axis = 0.75
    )
text(
  -0.7, 
  1,
  paste(
    "R^2 =", 
    signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
  )
)

@@

>>>>

```{r}
set.seed(1)
n_spp = 250
n_loc = 2500
n_env = 5
n_gibbs = 1000

pair_type = sample.int(
  2, 
  size = choose(n_spp, 2), 
  replace = TRUE,
  prob = c(.1, .9)
)

means = numeric(length(pair_type))
means[pair_type == 1] = 0
means[pair_type == 2] = 0

sds = numeric(length(pair_type))
sds[pair_type == 1] = 1.0
sds[pair_type == 2] = 0.2

true_beta_vec = rnorm(
  choose(n_spp, 2),
  mean = means, 
  sd = sds
)

plot(density(true_beta_vec))

@@

true_env = matrix(rnorm(n_loc * n_env, 0, 2), nrow = n_loc)

true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 1), nrow = n_env)

true_alpha_species = rnorm(n_spp, -2)

true_alpha = true_env %*% true_alpha_env %plus% true_alpha_species
```

@@

```{r}
# Are there lurking factors of two because I use the whole symmetric matrix?
beta_prior = rosalia::make_logistic_prior(scale = 0.1)$log_grad
alpha_env_prior = rosalia::make_logistic_prior(scale = 1)$log_grad 
alpha_species_prior = rosalia::make_logistic_prior(scale = 1)$log_grad
```

```{r}
initial_learning_rate = 1
momentum = 0.9
maxit = 5E3

pb <- progress_bar$new(
  format = "  Fitting [:bar] :percent eta: :eta",
  total = maxit,
  clear = FALSE
)

mses = numeric(maxit)
times = integer(maxit)

for(i in 1:maxit){
  pb$tick()

@@

  # Update learning rate
  learning_rate = initial_learning_rate * 500 / (499 + i)

@@

  # Calculate the gradient with respect to alpha and beta
  stats_difference = y_stats - y_sim_stats
  beta_grad = (stats_difference + beta_prior(beta)) / n_loc

  alpha_species_grad = (diag(stats_difference) + alpha_species_prior(alpha_species)) / n_loc
  diag(beta_grad) = 0
  y_env_difference = y_env_stats - y_sim_env_stats
  alpha_env_grad = (y_env_difference + alpha_env_prior(alpha_env))  / n_loc

  # Calculate parameter updates
  delta_beta = beta_grad * learning_rate / 10 + 
    momentum * delta_beta
  delta_alpha_species = alpha_species_grad * learning_rate + 
    momentum  * delta_alpha_species
  delta_alpha_env = alpha_env_grad * learning_rate + 
    momentum  * delta_alpha_env

@@

  mses[i] = mean((true_beta - beta)^2)
  times[i] = as.integer(Sys.time())

  if(i%%10 == 0){
    plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
         pch = 16, cex = .5, col = "#00000050",
         xlab = "Estimated interaction strength",
         ylab = "\"True\" interaction strength",
         main = paste(
           "R^2 =", 
           signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
         )
    )
    abline(v = 0, h = 0, a = 0, b = 1)
  }
}
```

```{r}
pdf("poster-beta.pdf", width = 4, height = 5)
plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
     pch = 16, cex = .4, col = "#00000020",
     xlab = "Estimated interaction strength",
     ylab = "\"True\" interaction strength",
     main = paste0("Recoverd pairwise interactions", "\nN = ", choose(n_spp, 2)),
     las = 1
    )
text(
  -0.7, 
  1,
  paste(
    "R^2 =", 
    signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
  )
)

@@

```{r}
covs =cov(y)
cor(covs[upper.tri(covs)], true_beta[upper.tri(true_beta)])^2
```

```{r}
library(rags2ridges)
inv = -ridgeS(cov(y), 5 / nrow(y))
cor(inv[upper.tri(inv)], true_beta[upper.tri(true_beta)])^2
```

```{r}
library(cowplot)
library(ggplot2)
out = plot_grid(
  ggplot(NULL, aes(x = beta[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#DDDDDD", high = "black", trans = "log10") +
    stat_smooth(method = "lm", se = FALSE, size = 1),

  ggplot(NULL, aes(x = inv[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#DDDDDD", high = "black", trans = "log10") +
    stat_smooth(method = "lm", se = FALSE, size = 1),

  ggplot(NULL, aes(x = covs[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#DDDDDD", high = "black", trans = "log10") +
    stat_smooth(method = "lm", se = FALSE, size = 1),
  nrow = 1,
  labels = c("A. Network estimates", "B. Partial covariance", "C. Covariance")
)
save_plot("estimates.pdf", out, base_aspect_ratio = 3)

```

```{r}
pdf("chapter-3/convergence.pdf", width = 10, height = 7)
par(mfrow = c(1, 2))
par(mar = c(5, 5, 4, 2) + .1)
plot(function(x){exp(-x * 10) * .15}, type = "l", axes = FALSE, ylab = "mean square deviation from MLE", xlab = "time", main = "A. Convergence to known MLE\n(small network)", yaxs = "i");axis(1, labels = FALSE);axis(2, seq(0, .16, .02))
text(.5, .1, "(illustration of\nexpected results)")
plot(mses, type = "l", axes = FALSE, xlab = "time (units to be added)", ylab = "mean squared deviation\nfrom \"true\" values",ylim = c(0, .15), main = "B. convergence to unknown MLE\n(large network)", yaxs = "i");axis(2);axis(1, labels = FALSE)
dev.off()
```
